---
title: "Heart Disease Analysis"
author: Jed P
date: March 11, 2019
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}

library(tidyverse)
library(scales)
library(dplyr)
library(corrplot)
library(ROCR)
library(caTools)
library(rpart)
library(naivebayes)


heart<-read.csv("E:/Documents/R/Data/projects/data/heart.csv",
                  header = TRUE,
                  na.strings = "",
                  stringsAsFactors = FALSE
)

```


```{r}
dim(heart) 
na_count <- sort(sapply(heart, function(x) sum(is.na(x))),
                 decreasing = TRUE)
na_count
glimpse(heart)
colnames(heart)
```

age, sex, cp (chest pain type), tresbps (resting blood pressure), chol, fbs (fasting blood sugar > 120 mg/dl, 1= true; 0 = false), 
restecg (resting ecg), thalach (max heart rate), exang (exercise induced), oldpeak, slope, ca (# of major vessels colored by flourosopy), 
thal (3 = normal, 6 = fixed defect, 7 = reversible defect), target


lets look at the correlation between the features
```{r}
cor_heart = cor(heart)
corrplot(cor_heart, method = "ellipse", type="upper")
```

thanks to this correlation plot, we can see that features trestbps, \
chol, fbs, and restecg are the least correlated features.
but just for completion's sake, we will use cholesterol as a metric in graphs to verify.


# data cleaning
```{r}
heart$sex <- as.factor(heart$sex)
heart$sex <- ifelse(heart$sex == "0", "female", "male")
heart$cp <-as.factor(heart$cp)
heart$target <- as.factor(heart$target)
heart$restecg <- as.factor(heart$restecg)
heart$slope <- as.factor(heart$slope)
heart$exang <- as.factor(heart$exang)


colnames(heart)[1] <- "age"
colnames(heart)[3] <- "chest_pain"
colnames(heart)[6] <- "fasting_blood_sugar"
heart$fasting_blood_sugar <- ifelse(heart$fasting_blood_sugar == "0", "under 120 mg", "over 120mg")
heart$fasting_blood_sugar <- as.factor(heart$fasting_blood_sugar)

summary(heart)
glimpse(heart)
```

## Target
First identify what the target feature looks like by itself
```{r}
rate_affected <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(position = "dodge2", color = "black") +
  geom_text(stat = 'count', aes(label = ..count.., vjust=-0.4), position = position_dodge(width=0.9)) +
  ggtitle("Heart Disease")

rate_affected + labs(x = "Unaffected vs Affected",
                     y = "Frequency"
                     )
```

change the target variable from 0~1 to have heart disease or absent
```{r}
heart$target <- ifelse(heart$target == "1", "Present", "Absent")
heart$target <- as.factor(heart$target)
```

## age
```{r}
ggplot(data = heart) +
  geom_bar(aes(x=age), fill = "blue") +
  facet_wrap(~ target) +
  ggtitle("Age Compared to Heart Disease")
```

## looking at affected and unaffected rates in measures of cholesterol
### sorted by chest pain type

```{r}
age_chol2 <- ggplot(data = heart) +
  geom_point(mapping = aes(x = heart$age, y = heart$chol, color = heart$sex), position = "jitter") +
  facet_wrap(heart$target ~ heart$chest_pain, nrow = 4, ncol = 4)

age_chol2 + labs(x = "Age", 
                 y = " Cholesterol", 
                 color = "Sex", 
                 title = "Levels of Cholesterol in mg", 
                 subtitle = "sorted by presence of heart disease, chest pain type", 
                 caption = "(data source: Kaggle Heart Disease UCI)"
                 )
```

### sorted by rest ecg
```{r}
age_chol3 <- ggplot(data = heart) +
  geom_point(mapping = aes(x = heart$age, y = heart$chol, color = heart$sex), position = "jitter") +
  facet_wrap(heart$target ~ heart$restecg, ncol = 3, nrow = 2)

age_chol3 + labs(x = "Age", 
                 y = " Cholesterol", 
                 color = "Sex", 
                 title = "Levels of Cholesterol in mg", 
                 subtitle = "sorted by presence of heart disease, restecg", 
                 caption = "(data source: Kaggle Heart Disease UCI)"
                 )
```

### sorted by fasting blood sugar over 120 mg
```{r}
age_chol4 <- ggplot(data = heart) +
  geom_point(mapping = aes(x = heart$age, y = heart$chol, color = heart$sex), position = "jitter") +
  facet_wrap(heart$target ~ heart$fasting_blood_sugar, ncol = 2, nrow = 2)

age_chol4 + labs(x = "Age", 
                 y = " Cholesterol", 
                 color = "Sex", 
                 title = "Levels of Cholesterol in mg", 
                 subtitle = "sorted by presence of heart disease, fasting blood sugar over 120 mg", 
                 caption = "(data source: Kaggle Heart Disease UCI)"
)

```
## Sex
```{r}
ggplot(heart, aes(x = sex, fill = target)) +
  geom_bar(position = "dodge2", color = "black") +
  geom_text(stat = 'count', aes(label = ..count.., vjust=-0.4), position = position_dodge(width=0.9)) +
  ggtitle("Gender vs Disease") 
```  
```{r}
male_hearts <- filter(heart, sex == "male")
female_hearts <- filter(heart, sex== "female")

ggplot(female_hearts, aes(x = target)) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill = target), stat = "count", color = "black") +
  geom_text(aes( label = scales::percent((..count..)/sum(..count..)),
                 y= ((..count..)/sum(..count..))), stat= "count", vjust = -0.4) +
  ggtitle("Percentage of Females") +
  xlab("Unaffected vs Affected") +
  ylab("% of Female Population")

ggplot(male_hearts, aes(x = target)) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill = target), stat = "count", color = "black") +
  geom_text(aes( label = scales::percent((..count..)/sum(..count..)),
                 y= ((..count..)/sum(..count..))), stat= "count", vjust = -0.4) +
  ggtitle("Percentage of Males") +
  xlab("Unaffected vs Affected") +
  ylab("% of Male Population")
```



```{r}
heart_disease_category <- c('Absent' = 'No Heart Disease',
                            'Present' = 'Heart Disease')

thalsex <- qplot(x = thalach, data = heart, binwidth = 5, geom = 'freqpoly', color = sex) +
  scale_x_continuous(breaks = seq(0, 303, 10)) +
  facet_wrap(heart$target, nrow = 2, labeller = as_labeller(heart_disease_category))


thalsex + labs(x = "Thalach",
               y = "Thalach Frequency",
               color = "Sex",
               title = "Thalach Levels by Sex",
               subtitle = "max heart rate",
               caption ="(data source: Kaggle Heart Disease UCI)"
                )
```

Features more likely to affect presence of heart disease so far:
chest pain, age, sex, exang, old peak, ca, slope, thal, thalach

# Creating a model
```{r}
set.seed(999)
splitter = sample.split(heart$target, SplitRatio = 0.75)
train = subset(heart, splitter == TRUE)
test = subset(heart, splitter == FALSE)


glm_mod <- glm(target ~ sex + chest_pain + thalach + thal + ca + slope + exang + oldpeak, data=train, family = "binomial")
pred <- predict(glm_mod, newdata = train, type = "response")


summary(glm_mod)

ROCRpred <- prediction(pred, train$target)
plot(performance(ROCRpred,"tpr","fpr"), colorize=TRUE)

```
